I am extremely thankful for the continuous support provided by my supervisors, Professor Andrew Stewart and Professor Caroline Jay. Andrew’s mentorship during my BSc and MRes degrees was foundational in motivating my PhD, and he has generously offered his time, expertise, and his fresh outlook, throughout the entire process. Caroline's perspective has also been extremely valuable and her advice and wisdom have been instrumental in improving my work. I am very lucky to have been given the opportunity to collaborate with such a dedicated team. I am also indebted to Dr. Jen McBride for the considerate and pragmatic guidance which she offered as my advisor.

For productive conversations, practical assistance, and camaraderie, thanks to Libby Le-luan, Gabe Strain, and Paul Stott. Thanks also to all in room 124e, and to Tristan Johnston-Wood and Sam Langton, for sharing experiences of the PhD process. I am grateful to Gabe Strain, Boshuo Zhang, and Harvey Schneider for their hard work in helping to develop experimental stimuli during their Master’s programmes.

I would also like to thank my family and friends; in particular Kirsten, Debbie, Brendan, and Jack for their support and endless hospitality. My parents have contributed immensely in helping me to keep going and have never failed to offer insightful guidance or an opportunity to discuss my work. I am extremely grateful for their unwavering support.

Finally, very special thanks to Hannah, for being a constant source of positivity, and for her thoughtful advice, assistance, encouragement, and understanding.