This thesis has been submitted in journal format, as the research in each empirical chapter (Chapters 4-6) addresses a different aspect of the topic under investigation. These chapters are therefore presented as stand-alone articles\textbf{, and are written using first-person plural pronouns}. A preface and postscript accompany each of these chapters to contextualise the work within the thesis. Introductory chapters (Chapters 1-3) provide background covering all three empirical investigations, and the final chapter (Chapter 7) provides a synthesis of this work.

\begin{itemize} 
\item The content of Chapter 4 is under review at \textit{Transactions on Visualization and Computer Graphics}. I am the paper’s first author; I devised the experiments, performed data analysis, and wrote the first draft. Experimental materials were developed in collaboration with Gabriel Strain, who also provided comments and enhancements regarding experimental design and report writing. The project was supervised by Caroline Jay and Andrew Stewart, who also provided comments and enhancements regarding experimental design, data analysis, and report writing.
\item The content of Chapter 5 has been published in \textit{Behaviour and Information Technology}. I am the paper’s first author; I devised the experiment, performed data analysis, and wrote the first draft. Experimental materials were developed in collaboration with Boshuo Zhang. The project was supervised by Caroline Jay and Andrew Stewart, who also provided comments and enhancements regarding experimental design, data analysis, and report writing.
\item The content of Chapter 6 is in preparation for submission to \textit{Transactions in Visualization and Computer Graphics}. I am the paper’s first author; I devised the experiments, performed data analysis, and wrote the first draft. Experimental materials were developed in collaboration with Harvey Schneider. The project was supervised by Caroline Jay and Andrew Stewart, who also provided comments and enhancements regarding experimental design, data analysis, and report writing.\\
\end{itemize} 